---
title: "Ashar Kurnia_Week 3"
author: "ashar"
date: "2026-02-18"
output: pdf_document
---

# A. Introduction
	Dengue virus (DENV) infects hundreds of millions of people annually, yet there is only a limited knowledge of the host immune response to dengue. Here, we used a systems biological approach to perform a detailed analysis of the innate immune response to DENV infection in the whole blood samples of acutely infected humans in Bangkok, Thailand.
	
	
	
# B. Method
#1. Package installation
```{r}
#a. Install BiocManager (manajer paket Bioconductor) 
if (!require("BiocManager", quietly = TRUE))  {
  install.packages("BiocManager") 
}

#b. Install paket Bioconductor (GEOquery & limma) 

BiocManager::install(c("GEOquery", "limma"), force = T) 

#c. Install annotation package platform
#GPL96 = Affymetrix Human Genome U133A 
BiocManager::install("hgu133a.db", ask = FALSE, update = FALSE)



#d. Install cRAN package for Visualization and Manipulation Data

install.packages(c("pheatmap", "ggplot2", "dplyr"))
if (!requireNamespace("umap", quietly = TRUE)) {
  install.packages("umap")
}


```

#2. Library calling
```{r}
library(GEOquery)
library(limma)
library(pheatmap)
library(ggplot2)
library(dplyr)
library(hgu133a.db)
library(AnnotationDbi)
library(umap)
```
#3. Get Datasets for GEO (GSE51808 datasets are used in this study)
```{r}
# The datasets contain 28 samples in tnfection time point, 19 in convalescent time point, 9 healthy control
options(timeout = 1000)
gset <- getGEO("GSE51808", GSEMatrix = TRUE, AnnotGPL = TRUE)[[1]] 

```
#4. Pre-Proccessing Gene Expression
```{r}
#exprs(): get expression gene matrix
ex <- exprs(gset)
#quantile(): count quantil value
qx <- as.numeric(quantile(ex, c(0, 0.25, 0.5, 0.75, 0.99, 1), na.rm = FALSE))
#LogTransform
LogTransform <- (qx[5] > 100) |  (qx[6] - qx[1] > 50 && qx[2] > 0)

#IF statement:
#If LogTransform = TRUE, run log2
LogTransform <- NA
if (isTRUE(LogTransform)) {
    print("Transforming")
}
```
#5. Group Sample Identification
```{r}
group_info <- pData(gset)[["source_name_ch1"]]
print(group_info)
#make.names(): convert text into valid R format
groups <- make.names(group_info)
print(groups)
#factor():
#Convert categoric data into factor
#Factor is very important for statistic in R
gset$groups <- factor(groups)
table(gset$group)  # Replication check for ebayes analysis (datasets must have replication)


#levels(): see category in a factor
group_name <- levels(gset$group)

#Merge group
group_name <- as.character(gset$group)

# If contain dengue word→ dengue
group_name[grepl("dengue", group_name, ignore.case = TRUE)] <- "dengue"

# if contain normal and control → normal
group_name[grepl("normal|control|healthy", group_name, ignore.case = TRUE)] <- "normal"

# Convert to factor
gset$group <- factor(group_name)

# Result
levels(gset$group)
table(gset$group)
```
# 6. Matrix Design
```{r}
design <- model.matrix(~0 + gset$group)
colnames(design) <- c("dengue", "normal")


# Make contrast
contrast.matrix <- makeContrasts(
  dengue_vs_normal = dengue - normal,
  levels = design
)

contrast.matrix
```


# 7. Differential Gene Expression Analysis (LIMMA)
```{r}
#Build linear model for gene expression

# linear model
fit <- lmFit(exprs(gset), design)



#contrasts.fit(): apply contrast to model

fit2 <- contrasts.fit(fit, contrast.matrix)

sum(is.na(ex))
sum(is.infinite(ex))

#eBayes():
#Empirical Bayes for stabilize varians model
fit2 <- eBayes(fit2)

#topTable():
#Get final result of DEG
#adjust = "fdr" -> multiple testing correction
#p.value = 0.01  -> define significant gene
topTableResults <- topTable(
  fit2,
  adjust = "fdr",
  sort.by = "B",
  number = Inf,
  p.value = 0.01
)

head(topTableResults)



```

# 8.UMAP Visualization
```{r}
umap_input <- t(ex)

#Run UMAP
umap_result <- umap(umap_input)

# Save to dataframe
umap_df <- data.frame(
  UMAP1 = umap_result$layout[, 1],
  UMAP2 = umap_result$layout[, 2],
  Group = gset$group
)

#UMAP Plot
ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = Group)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "UMAP Plot Sampel Berdasarkan Ekspresi Gen",
    x = "UMAP 1",
    y = "UMAP 2"
  )

```

#9. Volcano plot
```{r}
volcano_data <- data.frame(
  logFC = topTableResults$logFC,
  adj.P.Val = topTableResults$adj.P.Val,
  Gene = rownames(topTableResults)
)

#Status gene classification
volcano_data$status <- "NO"
volcano_data$status[volcano_data$logFC > 1 & volcano_data$adj.P.Val < 0.01] <- "UP"
volcano_data$status[volcano_data$logFC < -1 & volcano_data$adj.P.Val < 0.01] <- "DOWN"


#Visuazlization
volcano_plot <- ggplot(volcano_data, aes(x = logFC, y = -log10(adj.P.Val), color = status)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(values = c("DOWN" = "blue", "NO" = "grey", "UP" = "red")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  geom_hline(yintercept = -log10(0.01), linetype = "dashed") +
  theme_minimal() +
  ggtitle("Volcano Plot DEG Dengue")

```

#10. Heatmap Visualization
```{r}
# expression matrix
mat <- exprs(gset)

# Get fit gene
common_genes <- intersect(rownames(topTableResults), rownames(mat))

# Get top 50
top50 <- topTableResults[common_genes, ]
top50 <- top50[order(top50$adj.P.Val), ]
top50 <- top50[1:50, ]

# Heatmap matrix
mat_heatmap <- mat[rownames(top50), ]

# Remove NA
mat_heatmap <- mat_heatmap[complete.cases(mat_heatmap), ]

# annotation
annotation_col <- data.frame(Group = gset$group)
rownames(annotation_col) <- colnames(mat_heatmap)


```


# 11. Save work

```{r}
write.csv(topTableResults, "Hasil_GSE10072_DEG.csv")

message("Analysis is done. File are saved.")
```

# C. Result and Interpretation
```{r}

#1. Volcano plot
volcano_plot
```

The volcano plot shows of abundance downregulated gene. We can conclude that Dengue might affect in host immune gene expression.



```{r}
#2.Heatmap
pheatmap(
  mat_heatmap,
  scale = "row",                 # Z-score per gen
  annotation_col = annotation_col,
  show_colnames = FALSE,         # nama sampel dimatikan
  show_rownames = TRUE,
  fontsize_row = 7,
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  clustering_method = "complete",
  main = "Top 50 Differentially Expressed Genes"
)
```
The heatmap shows larger number of downregulated gene in Dengue group which is complement with volcano plot.

# D. Conclusion
The abundance of dowregulated gene shows that Dengue fever is significantly affect host immune gene expression

tinytex::install_tinytex()
